High-Dimensional Regularized Discriminant Analysis

### 2013-12-16 reports/paper-hdrda.Rnw
The intro's layout is closer to where it needs to be.
- Intro still needs polish.
- Also, slightly reorganized paper's layout.
- Related to #10 and #15.

Linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) are
classical approaches to supervised learning and are widely used in a variety of
areas, such as medical research, facial recognition, and TODO
\cite{Dudoit:2002ev,TODO1,TODO2} Both the LDA and QDA classifiers are based on
the assumption that the underlying distribution of each of the $K$ classes are
multivariate normally distributed. While both methods are popular in practice,
they often cannot be applied to modern data sets because the sample covariance
matrices employed are singular when the number of features exceeds the total
sample size $N$. For instance, standard benchmark microarray data sets used in
biological research often consist of between $p = 10,000$ and $55,000$ probes
and fewer than $N = 100$ observations, in which case LDA and QDA are
incalculable.

\cite{Friedman:1989tm} proposed regularized discriminant analysis (RDA), which
partially pools the individual sample covariance matrices used in the QDA
classifier with the pooled sample covariance matrix from the LDA
classifier. After pooling the covariance matrices, the RDA classifier employs a
covariance-matrix regularization technique that shrinks the zero eigenvalues of
the resulting covariance matrices to ensure they are nonsingular. Similarly,
several researchers, including \cite*{Zhang:2010va}, \cite{Ye:2009gd},
\cite{Ji:2008wp}, \cite*{Guo:2007te}, \cite*{Srivastava:2007ud},
\cite{Ye:2006vx}, and \cite{Ye:2005uu}, have proposed regularization methods to
improve the estimation of the population covariance matrices before applying
either the LDA or QDA classifiers. Covariance-matrix regularization methods
often are based on the well-known ridge-regression approach of
\cite{Hoerl:1970cd} and adjust the eigenvalues of the sample covariance matrices
to ensure positive definiteness. For excellent overviews of the regularization
problem in discriminant analysis, see \cite{Murphy:2012uq},
\cite{Mkhadri:1997gy}, and \cite{Friedman:1989tm}.

% TODO: Should we remove this paragraph? Keep a couple of the recent references
% though.
Alternatively, many researchers, including \cite{Mai:2012bf}, \cite{Fan:2012iq},
\cite{Merchante:2012vk}, \cite{Clemmensen:2011kr}, and \cite{Huang:2010ju}, have
introduced restrictions on the population covariance matrices in linear and
quadratic discriminant analysis to reduce the number of covariance parameters
that must be estimated to ensure that the resulting classifiers are
calculable. For instance, \cite{Dudoit:2002ev} have modified the linear and
quadratic discriminant analysis classifiers by assuming that the population
covariance matrices are diagonal, resulting in classifiers that are special
cases of the Naive Bayes classifier \citep[Chapter 4]{Hastie:2008dt}.

We reformulate the RDA classifier. Our parameterizaton yields 1) improved
interpretation, 2) nice properties that facilitate dimension reduction, and 3)
as a result of the dimension reduction, efficient calculation for
high-dimensional data.

Here, we propose the high-dimensional regularized discriminant analysis
(\emph{HDRDA}) classifier by first constructing a pooling matrix as a convex
combination of the covariance-matrix estimators employed in linear and quadratic
discriminant analysis. Given that the assumption of equal population covariance
matrices for linear discriminant analysis is often too stringent in practice
\citep{Clemmensen:2011kr}, our proposed classifier can relax the linearity
assumption by instead incorporating unequal covariance matrices while retaining
linear and quadratic discriminant analysis as special cases. We then establish
the \emph{HDRDA} classifier by applying covariance-matrix regularization with a
parameterization that includes a large family of shrinkage methods, such as
those from \cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods
studied by \cite{Ramey:2013ji} and \cite{Xu:2009fl}. This regularization has the
effect of stabilizing the eigenvalues of the covariance-matrix estimators and
ensuring that the classifier is well-posed.

We show that our proposed classifier is invariant to any adjustments of the
eigenvalues corresponding to the null space of the pooled sample covariance
matrix. Applying a reasoning similar to \cite{Ye:2006tq}, we decompose the
decision rule of the \emph{HDRDA} classifier and show that the sum related to
the null space of the pooled sample covariance matrix can be discarded. Hence,
no gain in classification performance can be obtained by adjusting these zero
eigenvalues. This invariance property provides a substantial increase in
computational efficiency of model training and selection. In fact, we show that
the inverse and determinants can be reduced to calculations of $n_k \times n_k$
matrices, where $n_k$ is the sample size of the $k$th class, $k = 1, \ldots, K$.
This reduction in calculation is clearly advantageous because $n_k$ is often
quite small in microarray data sets (e.g., 10--20 observations).

@@

\subsection{Preliminaries}
To facilitate our discussion of covariance-matrix regularization and regularized
discriminant analysis, we require the following notation. Let $\mathbb{R}_{a
  \times b}$ denotes the matrix space of all $a \times b$ matrices over the real
field $\mathbb{R}$. Denote by $\bm I_m$ the $m \times m$ identity matrix, and
let $\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm
0_m$ is understood to denote $\bm 0_{m \times m}$. Let $\bm A \oplus \bm B$
denote the direct sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in
\mathbb{R}_{s \times s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^{T}$,
$\bm A^+$, and $\mathcal{N}(\bm A) = \{\bm z \in \mathbb{R}_{p \times 1} : \bm A
\bm z = \bm 0_{m \times 1}\}$ denote the transpose, the Moore-Penrose
pseudoinverse, and the null space of $\bm A \in \mathbb{R}_{m \times p}$,
respectively. Let $\mathbb{R}_{p \times p}^{>}$ denote the cone of real $p
\times p$ positive-definite matrices. Similarly, let $\mathbb{R}_{p \times
  p}^{\ge}$ denote the cone of real $p \times p$ positive-semidefinite
matrices. Let $V^{\perp} = \{\bm y \in \mathbb{R}_{p \times 1} : \bm y^{T} \bm v
= 0$ for all $\bm v \in V \}$ denote the orthogonal complement of a vector space
$V \subset \mathbb{R}_{p \times 1}$. For $c \in \mathbb{R}$, let $c^+ = 1/c$ if
$c \ne 0$ and $0$ otherwise.

@@

\subsection{Model Specification}

@@

\subsection{Relationship with Regularized Discriminant Analysis}

@@

To further improve the estimation of $\bm \Sigma_k$ and to stabilize the inverse
of \eqref{eq:sig-lambda}, \cite{Friedman:1989tm} has proposed the biased
covariance-matrix estimator
\begin{align}
          \widehat{\bm \Sigma}_k(\lambda, \gamma) = (1 - \gamma) \widehat{\bm
            \Sigma}_k(\lambda) + \gamma \frac{\tr\{\widehat{\bm
              \Sigma}_k(\lambda)\}}{p} \bm I_p,\label{eq:sig-rda}
          \end{align}
where $\gamma \in [0, 1]$ is a regularization parameter that controls the
shrinkage of \eqref{eq:sig-rda} towards $\bm I_p$ and is weighted by the average
of the eigenvalues of \eqref{eq:sig-lambda}. Finally, the RDA classifier is
established by substituting \eqref{eq:sig-rda} into \eqref{eq:qda}.

\section{Properties of the HDRDA}

\subsection{Enhanced Computational Efficiency}
\label{sec:computational-efficiency}

@@

>>>>

Discriminant analysis is a widely used technique to classify an unlabeled
$p$-dimensional vector $\bm x$ to one of $K$ unique, known classes by
constructing a classifier from $N$ training observations. This is useful in a
variety of areas, such as medical research, facial recognition, and TODO
\cite{Dudoit:2002ev,TODO1,TODO2} because data-driven decision are becoming
increasingly common.  Although a large number of classification algorithms have
been proposed, linear discriminant analysis (LDA) and quadratic discriminant
analysis (QDA) remain popular in practice because of their simplicity, speed of
computation, and interpretability. Both the LDA and QDA classifiers are based on
the assumption that each of the $K$ classes is multivariate normally
distributed. While both methods are popular in practice, they often cannot be
applied to modern data sets because the sample covariance matrices employed are
singular when the number of features exceeds the total sample size $N$. For
instance, standard benchmark microarray data sets used in biological research
often consist of between $p = 10,000$ and $55,000$ probes and fewer than $N =
100$ observations, in which case LDA and QDA are incalculable.\footnote{TODO:
  Bring back the curse of dimensionality reference.}

As a result, several researchers have applied covariance-matrix regularization
to the LDA and QDA classifiers. Covariance-matrix regularization methods often
are based on the well-known ridge-regression approach of \cite{Hoerl:1970cd} and
adjust the eigenvalues of the sample covariance matrices to ensure positive
definiteness.  \cite*{Zhang:2010va}, \cite{Ye:2009gd}, \cite{Ji:2008wp},
\cite*{Guo:2007te}, \cite*{Srivastava:2007ud}, \cite{Ye:2006vx}, and
\cite{Ye:2005uu}, have proposed regularization methods to improve the estimation
of the population covariance matrices before applying either the LDA or QDA
classifiers. Alternatively, many researchers, including \cite{Mai:2012bf},
\cite{Fan:2012iq}, \cite{Merchante:2012vk}, \cite{Clemmensen:2011kr}, and
\cite{Huang:2010ju}, have introduced restrictions on the population covariance
matrices to reduce the number of parameters that must be estimated. For
instance, \cite{Dudoit:2002ev} have modified the linear and quadratic
discriminant analysis classifiers by assuming that the population covariance
matrices are diagonal, resulting in classifiers that are special cases of the
Naive Bayes classifier \citep[Chapter 4]{Hastie:2008dt}. For excellent overviews
of the regularization problem in discriminant analysis, see \cite{Murphy:2012uq}
and \cite{Mkhadri:1997gy}.

In a classic paper, \cite{Friedman:1989tm} proposed regularized discriminant
analysis (RDA), which partially pools the individual sample covariance matrices
used in the QDA classifier with the pooled sample covariance matrix from the LDA
classifier. After pooling the covariance matrices, the RDA classifier employs a
covariance-matrix regularization technique that shrinks the zero eigenvalues of
the resulting covariance matrices to ensure they are nonsingular. Given that the
assumption of equal population covariance matrices for the LDA classifier is
often too stringent in practice \citep{Clemmensen:2011kr}, the RDA classifier
can relax the linearity assumption by instead incorporating unequal covariance
matrices while retaining as special cases the LDA and QDA classifiers. Despite
its success, the RDA classifier is often impractical because the model-selection
process for its two tuning parameters requires excessive computation. Moreover,
the pooling parameter has no clear intrepretation, and others
complaints\footnote{TODO: Add some more complaints.}\footnote{TODO: Citations
  needed.}.

We consider the RDA classifier in an attempt to study rigorously some of its
properties. Similar to \cite{Hastie:2008dt} and \cite{Halbe:2007ea}, we first
reparameterize the RDA classifier's covariance-matrix estimator as a convex
combination of the covariance-matrix estimators employed in the LDA and QDA
classifiers. Our parameterizaton yields 1) improved interpretation, 2) nice
properties that facilitate dimension reduction, and 3) as a result of the
dimension reduction, efficient calculation for high-dimensional data. We then
establish the high-dimensional RDA (\emph{HDRDA}) classifier by applying
covariance-matrix regularization with a parameterization that includes a large
family of shrinkage methods, such as those from \cite{Srivastava:2007ww},
\cite{Rao:1971ul}, and other methods studied by \cite{Ramey:2013ji} and
\cite{Xu:2009fl}. This regularization has the effect of stabilizing the
eigenvalues of the covariance-matrix estimators and ensuring that the classifier
is well-posed.

We show that our proposed classifier is invariant to any adjustments of the
eigenvalues corresponding to the null space of the pooled sample covariance
matrix. Applying a reasoning similar to \cite{Ye:2006tq}, we decompose the
decision rule of the \emph{HDRDA} classifier and show that the sum related to
the null space of the pooled sample covariance matrix can be discarded. Hence,
no gain in classification performance can be obtained by adjusting these zero
eigenvalues. This invariance property provides a substantial increase in
computational efficiency of model training and selection. In fact, we show that
the inverse and determinants can be reduced to calculations of $n_k \times n_k$
matrices, where $n_k$ is the sample size of the $k$th class, $k = 1, \ldots, K$.
This reduction in calculation is clearly advantageous because $n_k$ can be small
in modern high-dimensional data sets (e.g., 10--20 observations in
gene-expression microarray data).

@@

\section{Preliminaries}
To facilitate our discussion of covariance-matrix regularization and regularized
discriminant analysis, we require the following notation. Let $\mathbb{R}_{a
  \times b}$ denotes the matrix space of all $a \times b$ matrices over the real
field $\mathbb{R}$. Denote by $\bm I_m$ the $m \times m$ identity matrix, and
let $\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm
0_m$ is understood to denote $\bm 0_{m \times m}$. Let $\bm A \oplus \bm B$
denote the direct sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in
\mathbb{R}_{s \times s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^{T}$,
$\bm A^+$, and $\mathcal{N}(\bm A) = \{\bm z \in \mathbb{R}_{p \times 1} : \bm A
\bm z = \bm 0_{m \times 1}\}$ denote the transpose, the Moore-Penrose
pseudoinverse, and the null space of $\bm A \in \mathbb{R}_{m \times p}$,
respectively. Let $\mathbb{R}_{p \times p}^{>}$ denote the cone of real $p
\times p$ positive-definite matrices. Similarly, let $\mathbb{R}_{p \times
  p}^{\ge}$ denote the cone of real $p \times p$ positive-semidefinite
matrices. Let $V^{\perp} = \{\bm y \in \mathbb{R}_{p \times 1} : \bm y^{T} \bm v
= 0$ for all $\bm v \in V \}$ denote the orthogonal complement of a vector space
$V \subset \mathbb{R}_{p \times 1}$. For $c \in \mathbb{R}$, let $c^+ = 1/c$ if
$c \ne 0$ and $0$ otherwise.

@@

@@

\subsection{Relationship with Friedman's RDA Classifier}

@@

To further improve the estimation of $\bm \Sigma_k$ and to stabilize the inverse
of \eqref{eq:sig-lambda}, \cite{Friedman:1989tm} has proposed the biased
covariance-matrix estimator
\begin{align}
          \widehat{\bm \Sigma}_k(\lambda, \gamma) = (1 - \gamma) \widehat{\bm
            \Sigma}_k(\lambda) + \gamma \frac{\tr\{\widehat{\bm
              \Sigma}_k(\lambda)\}}{p} \bm I_p,\label{eq:sig-rda}
          \end{align}
where $\gamma \in [0, 1]$ is a regularization parameter that controls the
shrinkage of \eqref{eq:sig-rda} towards $\bm I_p$ and is weighted by the average
of the eigenvalues of \eqref{eq:sig-lambda}. Finally, the RDA classifier is
established by substituting \eqref{eq:sig-rda} into
\eqref{eq:qda}.\footnote{TODO: Discuss $\alpha_k \equiv 1-\gamma$, etc.}

\subsection{Properties of the HDRDA Classifier}
\label{sec:computational-efficiency}

@@

\subsection{Computationally Efficient Model Selection}
